Recognizing Strong Gravitational Lenses

### 2015-03-15 doc/Course_Paper/paper.tex
updating course paper

%%%%%%%%% ABSTRACT
\begin{abstract}
    The detection of strong gravitational lenses
\end{abstract}

@@

A consequence of Einstein's Theory of General Relativity is that mass bends the
path of light. Most of the time the deflections are very small; the original
'gravitational lens' that tested the veracity of Einstein's theory in the years
after the first world war was the sun, which deflected light from stars behind
it only a few seconds of arc. However, when light passes through a particularly
deep gravitational potential (say, near the center of the dark matter halo of a
galaxy cluster), the deflections can be particularly large, resulting in
brilliant arcs and multiple images. These strong deflections due to light
passing through a deep gravitational potential are termed strong gravitational
lenses.

@@

\begin{figure*}[!ht]
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/0.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/1.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/2.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/3.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
  \caption{Typical Space Warps duds. Insets indicate regions where volunteers typically
  clicked.}
  \label{fig:duds}
\end{figure*}

\begin{figure*}[!ht]
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/4.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/5.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/6.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/7.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
  \caption{Typical sims. Insets indicate the location of the lens in the
  image.}
  \label{fig:sims}
\end{figure*}

@@

We would like to examine the following questions:
\begin{itemize}
\item{ Do we have enough data to reasonably train and test a CNN? Can we get
       around this by artificially inflating the data, e.g. by adding rotated
     images?}
\item{ What processing needs to be done on the data? What kind of scaling of
       pixel data is appropriate for automated detection? Should we compress
     the five different 'colors' to a smaller number of dimensions?}
   \item{ How do citizen-scientists do compared with this automated system?}
   \item{Can we use the results of citizen-scientists to train the CNN?}
\item{What sets of classifications are needed? Are we better served sticking
       to 'lens' and 'not lens', or should we use several classification
       categories ('lensed arcs', 'lensed multiple images', 'non-lens pixel
     noise')?}
\end{itemize}

@@

We plan to train a classifier on this data using two different methods. First,
we will code our own convolutional net in python using \textsc{theano}. Second,
we will apply transfer learning techniques to train on a convolutional net
galaxy morphology classifier, which has graciously been made available to us by
Ryan Keisler and which achieved 7th place in the 2014 Galaxy Zoo Kaggle
competition.  This classifier runs $96\times96$ images through 3 convolutional
layers and 2 fully-connected layers and predicts a galaxy to have one of 37
enumerated morphologies. We plan to train our classifier on top of the first
fully-connected layer, which has 500 neurons.

@@

% TODO: Touch up description
\begin{figure}%[!ht]
\centering
\epsfig{file=Figures/architecture.pdf,width=\linewidth,angle=0,clip=}
\caption{Architecture of the convolution
  neural network trained on the galaxy zoo morphologies. The feature vector we
use comes from the fully connected layer.}
\label{fig:architecture}
\end{figure}

% TODO: Redescribe the network, including data augmentation and such.

@@

@@

%-------------------------------------------------------------------------
\section{Acknowledgements}
We received access to the dataset through Phil Marshall
(\url{pjm@slac.stanford.edu}), who also graciously explained to us how
\textsc{Space Warps}
currently works. The convolutional neural network upon which the transfer
learning is based was kindly provided to us by Ryan Keisler
(\url{rkeisler@stanford.edu}).
All errors in interpretation or otherwise are entirely our own.

@@

We have created the cutouts of our images from the catalogs.
We have begun constructing our own convolutional network with two
convolutional/max-pool layers and two fully-connected layers. Although we have
not yet trained on our images, we can load these images and run them
through our initialized classifier. Figures~\ref{fig:duds} and
\ref{fig:sims}\xspace show example fields with cutouts inlaid.

>>>>

%%%%%%%%% ABSTRACT
\begin{abstract}
    The detection of a large, representative set of strong gravitational lenses
    could greatly aid in our understanding of cosmology. Unfortunately they are
    quite rare, and the best techniques now revolve around squads of scientists
    manually scanning through images. This is presently borderline
    unsustainable and will be laughably inefficient with the advent of the
    Large Synoptic Survey Telescope. Here we examine the effectiveness of
    convolution neural networks and transfer learning for automated detection
    algorithms of strong gravitational lenses. We use images from the
    \textsc{Space Warps} project, a citizen science initiative to examine tens
    of thousands of fields of galaxies for the presence of strong gravitational
    lenses. 
    % TODO: Insert results!
    Scripts used in the analysis of this paper are freely available at
    \url{https://github.com/cpadavis/strongcnn}. The images are currently only
    available to those who contact the author, but will be available to the
    public in the near future.
\end{abstract}

@@

% TODO: insert references!

A consequence of Einstein's Theory of General Relativity is that mass bends the
path of light. Most of the time the deflections are very small; the original
'gravitational lens' that tested the veracity of Einstein's theory in the years
after the first world war was the sun, which deflected light from stars behind
it only a few seconds of arc. However, when light passes through a particularly
deep gravitational potential (say, near the center of the dark matter halo of a
galaxy cluster), the deflections can be particularly large, resulting in
brilliant arcs and multiple images. These strong deflections due to light
passing through a deep gravitational potential are termed strong gravitational
lenses. The gravitational lensing signal can heuristically be thought of as a
trade off between a couple factors: larger gradients in the gravitational
potential create larger distortions (and larger gradients in the gravitational
potential tend to reside near to the center of the foreground galaxy or galaxy
cluster); the more separated the foreground and background objects are, the
bigger the proportional distortion on the background object by the foreground
(but the farther away the background object is, the smaller it
appears\footnote{Note that this is only true in the ``low redshift'' universe:
  when objects are farther away than a cosmological redshift of $z \approx 2.5$
or about 2.5 Gyr after the birth of the Universe, they will actually
\textit{grow} in angular extent.}).

@@

\begin{figure*}[!ht]
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/4.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/5.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/6.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/7.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
  \caption{Typical sims. Insets indicate the location of the lens in the
  image. These insets are fed into our training system.}
  \label{fig:sims}
\end{figure*}

\begin{figure*}[!ht]
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/0.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/1.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/2.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.24\linewidth}
\centering\epsfig{file=Figures/gallery/3.png,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
  \caption{Typical Space Warps duds. Insets indicate regions where volunteers
  typically clicked. These insets are then fed into our training system.}
  \label{fig:duds}
\end{figure*}

@@

We would like to examine the following questions:
\begin{itemize}
\item{ Do we have enough data to reasonably train and test a CNN? Can we get
       around this by artificially inflating the data, e.g. by adding rotated
     images?}
\item{ How do citizen-scientists do compared with this automated system?}
\item{ Can we use the results of citizen-scientists to train the CNN?}
\item{ How well does using features extracted from a convolution neural network
       trained on galaxy morphology perform when determining the presence of
     strong gravitational lenses?}
\end{itemize}

@@

We train a classifier on this data using two different methods. First,
we code our own convolutional net in python using \textsc{theano}. Second,
we apply transfer learning techniques to train on a convolutional net
galaxy morphology classifier, which has graciously been made available to us by
Ryan Keisler and which achieved 7th place in the 2014 Galaxy Zoo Kaggle
competition.  This classifier runs $96\times96$ images through 3 convolutional
layers and 2 fully-connected layers and predicts a galaxy to have one of 37
enumerated morphologies. (See Figure~\ref{fig:ryan_architecture}.) We
train classifiers on top of the first fully-connected layer, which has 500
neurons.

%-------------------------------------------------------------------------
\subsection{The \textsc{Space Warps} Catalog}

\textsc{Space Warps} is a web-based service that enables the discovery of
strong gravitational lenses in wide-field imaging surveys by large numbers of
people. Carefully produced color composite images are displayed to volunteers
via a classification interface which records their estimates of the positions
of candidate lensed features. Simulated lenses, and expert-classified
non-lenses, are inserted into the image stream at random intervals; this
training set is used to give the volunteers feedback on their performance, and
to estimate a dynamically-updated probability for any given image to contain a
lens. Low probability systems are retired from the site periodically,
concentrating the sample towards a set of candidates; this ``stage 1'' set is
then re-classified by the volunteers in a second refinement stage. This ``stage
2'' has a different set of training images, ones that are generally considered
`harder'. Most stage 1 simulated lenses are very obvious\footnote{Very bright
and blue quasars multiply-imaged around a small red galaxy, very bright,
separated, and full \href{http://en.wikipedia.org/wiki/Einstein_ring}{Einstein
rings.}}, while simulated lenses in stage 2 are often much more subtle.
\footnote{Dim multiply-imaged quasars of varying magnitude, dim and incomplete
Einstein rings located close to a galaxy.} Figures~\ref{fig:duds} and
\ref{fig:sims}\xspace show example stage 2 fields with cutouts inlaid. Notice
that while the first three images in Figure~\ref{fig:sims} are very clearly
strong gravitational lenses \footnote{For the neophyte: the first is a broken
  blue arc around a central red galaxy; the second is a blue arc around a
  central yellow galaxy; the third is a multiply-imaged blue quasar (images
appear above and below the central galaxy), the fourth is a broken dim arc
located behind a very bright foreground galaxy.}, the fourth is very difficult
to find.  Unfortunately, we would very much like to find these, because there
are many such systems and they contain important information about the mass
structures at the centers of galaxies.\footnote{The trade off is this: the rate
  of alignment between foreground and background objects decreases as one
  decreases the area around a foreground object, but the strength of strong
  gravitational distortions -- and the signal we can pull out from identifying
  such systems -- increases as one gets closer to the center of the foreground
  object. A yet further complication to this is that background objects are
  naturally fainter than foreground objects, \textit{and} the foreground
  objects with the highest gravitational potential (and hence the largest
distortions of background images) tend to also be the brightest objects. Both
these complications render the task even more difficult.}
Figure~\ref{fig:duds} highlights the difficulties of this task. Each `dud' has
features that conceivably look like strong gravitational lensing, but are in
actuality some other confounding effect: color gradients from variations in the
Point Spread Function between the different color bands, dust surrounding a
galaxy, galaxies that are actually in the same cluster, and chance alignments
of background galaxies and foreground stars.

The fields users observe are $440\times440$ size images, containing multiple
potential locations for strong gravitational lenses, although it is unlikely
that a field contains more than one strong gravitational lens. In order to
generate $96\times96$ images of lenses and non-lenses, we use the recorded
estimates of the positions of candidate lensed feature. More specifically, we
apply the \textsc{DBSCAN} clustering algorithm, which agglomeratively grows
clusters such that that are within a minimum distance and contain a minimum
number of samples. \textsc{DBSCAN} is a convenient choice of a clustering
algorithm because it has a well-defined way of rejecting outliers, which we
generically interpret as genuine ``mis-clicks'' on the part of users. We are
very generous in the definition of a cluster and only require two members within
100 pixels of each other to form a cluster.

The fear of noisy clusters is this: the real task of our techniques is to
distinguish strong gravitational lens systems from other configurations of
galaxies (for example, random alignments of galaxies). Noisy clicks end up
creating random cutouts of the field, slightly changing the task of our
classifier to distinguishing strong gravitational lens systems from random
cutouts from the field of a galaxy survey.  Inclusion of noise however ended up
not being an issue for stage 2: non-lenses, where the correct action on the
part of the user is to leave no marker, have a median of 28 markers in stage 2,
while simulated lenses (where the correct answer is to click at a specific
location) have a median of 180 markers. In stage 1, the simulated lenses have a
median of 80 markers, while the duds have a median of 3 markers (and a mean of
9.6). It may be the case that noise is injected in the stage 1 non-lens sample.

Overall, our base dataset has 24,177 images from stage 1, of which 5159 are of
simulated lenses, and 1876 images from stage 2, of which 151 are simulated
lenses. We also have 9030 classifications that stage 2 users made of images in
the CFHTLS survey where it is unknown whether they contain a lens or not. From
these classifications, a list of approximately 40 candidate strong
gravitational lensing objects have been found which will soon receive
spectroscopic follow-up to confirm whether they are strong gravitational
lensing systems or not.\footnote{Spectroscopy can yield precise redshifts of
  different objects in a field. This way, if different parts of a strong
  gravitational lens arc are at the same redshift, or if the multiply-imaged
quasars are, then one can confirm that we are really seeing such a system.} A
future project with this work would be to link the probabilities from the
\textsc{Space Warps} system with the probabilities obtained by a detection
algorithm.

@@

% TODO: Touch up description
\begin{figure}%[!ht]
\centering
\epsfig{file=Figures/architecture.pdf,width=\linewidth,angle=0,clip=}
\caption{Architecture of the convolution
  neural network trained on the galaxy zoo morphologies. The feature vector we
use comes from the fully connected layer.}
\label{fig:ryan_architecture}
\end{figure}

\begin{figure*}%[!ht] 
\begin{minipage}[b]{0.48\linewidth}
\centering\epsfig{file=Figures/roc_curve_stage_1.pdf,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
\begin{minipage}[b]{0.48\linewidth}
\centering\epsfig{file=Figures/roc_curve_stage_2.pdf,width=\linewidth,angle=0,clip=}
\end{minipage} \hfill
  \caption{Receiver Operating Curves the \textsc{Space Warps} system and
    different linear classifiers trained on feature vectors extracted from a
    convolution neural network originally used to determine galaxy
  morphologies. We find that of the linear classifiers on the feature vectors,
support vector machine classifications perform best on the test dataset,
however all the feature vectors perform worse than the users themselves. Note
that the $x$-axis stops at a false positive rate of 0.5, and the $y$-axis
begins at a true positive rate of 0.5.}
\label{fig:roc_tl} 
\end{figure*}

% TODO: Redescribe the network, including data augmentation and such.
Transfer Learning provides a mechanism by which

@@

% TODO: make figure with probability distributions between classifiers and sw

@@

%-------------------------------------------------------------------------
\section{Acknowledgements}
We received access to the dataset through Phil Marshall
\footnote{\url{pjm@slac.stanford.edu}}, who also graciously explained to us how
\textsc{Space Warps}
currently works. The convolutional neural network upon which the transfer
learning is based was kindly provided to us by Ryan Keisler
\footnote{\url{rkeisler@stanford.edu}}.
All errors in interpretation or otherwise are entirely our own.

@@

We have created the cutouts of our images from the catalogs.
We have begun constructing our own convolutional network with two
convolutional/max-pool layers and two fully-connected layers. Although we have
not yet trained on our images, we can load these images and run them
through our initialized classifier.
